$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-08-02 qdpm_paper.tex
Worked on text, updated bib

      height=5.0ex,% adjust to suit

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer ($n_0 = 1$), two dimers ($n_0 = 2$), and an even ($n_P = 0$) or odd ($n_P = 1$) number of dimers touching each vertex.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires only $Q_v \mod 2$ to vanish $\forall v$.

@@

We define the string operator is defined as
\begin{equation}
S_\gamma \left( \alpha \right) \equiv e^{i \alpha \left( \Phi_\gamma - \tilde{\Phi}_\gamma \right)} \label{eq:Sgamma}
\end{equation}
where $\alpha$ such that $S_\gamma$ commutes with $\HGQDM$ everywhere except at the ends of $\gamma$. Therefore, for closed loops $\Gamma$, $[ S_\Gamma,\HGQDM]=0$. For open strings, $S_\gamma$ does not commute with the Hamiltonian and can be viewed as creating a pair of ``magnetic charges'' at the ends of the string: correspondingly $S_\gamma \ket{\Phi_{\RK}}$ can used as a trial wavefunction for an excited state with a pair of magnetic charges.

@@

\begin{table}
    \begin{tabular}{| l | c | c |}
  	\hline
     & $c$ & $\bar{c}$ \\
	\hline
    $i$ & $\monomermvA$ &  $\monomermvB$  \\
    \hline
  \end{tabular}
  \caption{The pair of local dimer \& monomer configurations $c\leftrightarrow\bar{c}$ generate
  minimal local monomer dynamics. Monomers always remain on their origonal sublattice. }
  \label{tab:pentamerMoves}
\end{table}

@@

Here we discuss the dimer density correlations in the RK ground state of the QDPM. In \figref{fig:dimer_gry_scale_qdpm} we plot the bare dimer-dimer density correlation function $\langle n_{\bm{0}} n_{\bm{r}}\rangle$ on an $L=16$ lattice; here the origin $\bm{0}$ represents the blue link, and $\bm{r}$ varies over all other links of the lattice. We see no evidence of symmetry breaking in the dimer correlations on lattices up $L=64$. We will restrict the rest of our disucssion to correlations between horizontal of dimers, as we find no rotational symmetry breaking, or qualitative difference between the the correlations between parallel dimers and perpendicular dimers.

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_gry_scale_qdpm.pdf}
        \caption{({\it left}) Gray scale of the dimer-dimer correlation function $\langle n_0 n_{\bm{r}}\rangle$ on an $L=16$ lattice. The origin corresponds to the blue link, and $\bm{r}$ varies over all other links of the lattice. Red links have correlations that are off the given scale.~\note{Owen: Is this the bare or connected correlation function? What do these plots look like on a log scale?} ({\it right}) $Z_3$ magnetic string correlation function. }
        \label{fig:dimer_gry_scale_qdpm}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
   \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{qdpm_dmr_vis_HM_16x16.pdf}
        \caption{({\it left}) The horizontal dimer density structure factor $S_d(\bm{k})$ for the QDPM RK ground state on an $L=32$ lattice. The lack of $x\leftrightarrow y$ rotational symmetry is due to the fact that horizontal links have been isolated. ({\it right}) \note{Owen: What does this vison structure factor look like w/o the factor for the background dimerization?}}
        \label{fig:structure_factor} 
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{vis_dmr_qdpm_struc_fac_32x32.pdf}
        \caption{ The horizontal dimer-dimer structure factor $S_d(\bm{k})$ ({\it top}) and the ({\it bottom})
        shown along the path in $k$ space connecting the $(0,0)$, to $(\pi,0)$, $(\pi,\pi)$, $(0,0)$
        points for $\gamma=0.5$. Statistical errors are smaller than the symbols. \note{Owen: Make the k's in the y-axis bold. What do there look like if you added $L=16$ data to these plots?}}
        \label{fig:structure_factor_path}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%

@@

To investigate the $Z_3$ gauge theory description of the QDPM RK ground state, we study the $Z_3$ magnetic string correlation function as defined in \eqref{eq:Sgamma} \& \eqref{eq:SQDPM}:
\begin{equation}
C_M \left( \bm{r}_{pp'} \right) \equiv \left \langle \exp \left[ \frac{2 \pi}{3} i \left( \Phi_{\gamma_{pp'}} -\tilde{\Phi}_{\gamma_{pp'}} \right) \right] \right \rangle
\end{equation}
where $p$ and $p'$ are two plaquettes, and $\gamma_{pp'}$ is any open string on the dual lattice connecting $p$ and $p'$.

@@

>>>>

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer ($n_0 = 1$), two dimers ($n_0 = 2$), and an even ($n_P = 0$) or odd ($n_P = 1$) number of dimers touching each vertex.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires only $(Q_v\mod 2)$ to vanish.

@@

We define the string operator is defined as
\begin{equation}
S_\gamma \left( \alpha \right) \equiv e^{i \alpha \left( \Phi_\gamma - \tilde{\Phi}_\gamma \right)} \label{eq:Sgamma}
\end{equation}
where $\alpha$ such that $S_\gamma$ commutes with $\HGQDM$ everywhere except at the ends of $\gamma$. Therefore, for closed loops $\Gamma$, $[ S_\Gamma,\HGQDM]=0$. For open strings, $S_\gamma$ does not commute with the Hamiltonian and can be viewed as creating a pair of ``magnetic charges'' at the ends of the string: correspondingly $S_\gamma \ket{\Phi_{\RK}}$ can used as a trial wavefunction for an excited state with a pair of magnetic vortices. If a state has topological order with deconfined electric charges, these magnetic vortices should be gapped excitations, and correspondingly the magnetic string operator should decay exponentially~\cite{Read1989a,Senthil2000,Senthil2001e}. The condensation of magnetic vortices can drive a phase transition from a topologically ordered phase to a conventionally order phase~\cite{Jalabert1991,Ralko2007}. Thus, the magnetic string correlation can be used as test for deconfinement of electric charges.

@@

@@

Here we discuss the dimer density correlations in the RK ground state of the QDPM. In \figref{fig:dimer_vison_heatmap} we plot the bare dimer-dimer density correlation function $\langle n_{\bm{0}} n_{\bm{r}}\rangle$ on an $L=16$ lattice; here the origin $\bm{0}$ represents the blue link, and $\bm{r}$ varies over all other links of the lattice. We see no evidence of symmetry breaking in the dimer correlations on lattices up $L=64$. We will restrict the rest of our disucssion to correlations between horizontal of dimers, as we find no rotational symmetry breaking, or qualitative difference between the the correlations between parallel dimers and perpendicular dimers.

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_gry_scale_qdpm.pdf}
        \caption{({\it left}) Gray scale of the dimer-dimer correlation function $\langle n_{\bm{0}} n_{\bm{r}}\rangle$ on an $L=16$ lattice. The origin corresponds to the blue link, and $\bm{r}$ varies over all other links of the lattice. Red links have correlations that are off the given scale.~\note{Owen: Is this the bare or connected correlation function? What do these plots look like on a log scale?} ({\it right}) The $Z_3$ magnetic string correlation function $C_m(\bm{r}_{\bm{0}p})$, where the the correlations is with respect to the blue plaquette. \note{Are we using the real part? What does $\sqrt{C_M^*C_m}$ look like? Regardless we should make this a positive quantity so the lefthand side of the grayscale is zero.}}
        \label{fig:dimer_vison_heatmap}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
   \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{qdpm_dmr_vis_HM_16x16.pdf}
        \caption{({\it left}) The horizontal dimer density structure factor $S_d(\bm{k})$ for the QDPM RK ground state on an $L=32$ lattice. The lack of $x\leftrightarrow y$ rotational symmetry is due to the fact that horizontal links have been isolated. ({\it right}) \note{Owen: Is this the real part or $\sqrt{S_M^* S_M}$}}
        \label{fig:structure_factor} 
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{vis_dmr_qdpm_struc_fac_32x32.pdf}
        \caption{ The horizontal dimer-dimer structure factor $S_d(\bm{k})$ ({\it top}) and the ({\it bottom})
        shown along the path in $k$ space connecting the $(0,0)$, to $(\pi,0)$, $(\pi,\pi)$, $(0,0)$
        points for $\gamma=0.5$. Statistical errors are smaller than the symbols. \note{Owen: Make the k's in the y-axis bold. What do there look like if you added $L=16$ data to these plots?}}
        \label{fig:structure_factor_path}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%

@@

To investigate the $Z_3$ gauge theory description of the QDPM RK ground state, we study the $Z_3$ magnetic string correlation function as defined in \eqref{eq:Sgamma} \& \eqref{eq:SQDPM}:
\begin{equation}
C_M \left( \bm{r}_{pp'} \right) \equiv \left \langle \exp \left[ \frac{2 \pi}{3} i \left( \Phi_{\gamma_{pp'}} -\tilde{\Phi}_{\gamma_{pp'}} \right) \right] \right \rangle
\end{equation}
where $p$ and $p'$ are two plaquettes, and $\gamma_{pp'}$ is any open string on the dual lattice connecting $p$ and $p'$. \note{Something about the background dimerization and gauge fixing here. Are we using the real parts or $\sqrt{C_M^*C_M}$?}

In \figref{fig:dimer_vison_heatmap}, we have ploted $C_M$ for an $L=16$ lattice, which displays short distance correlations that quickly decay. \note{Comment here about gauge fixing breaking symmetry, if still present...}. We find no evidence of magnetic vortex condensation on lattices up to $L=64$.

@@

To study the confinement of electrically charged defects in the QDPM, we have Monte Carlo sampled the RK ground state with a single pair of monomers. We initialize the system with a monomer on each sub-lattice and supplement the local Monte Carlo updates with updates of the form:
\begin{equation}
\monomermvA  \Longleftrightarrow \monomermvB
\end{equation}
which appear to provide ergodic dynamics in the monomer-dimer-pentamer configuration space. Notice that monomers remain on the same sub-lattice, which is consistent with their assignment of a $Z_3$ electric charge of $\pm1$, depending on the sub-lattice.
